---
title: "K-Means and K-Medoids Clustering on Simulated Data"
author: "Halil Bisgin"
output: html_document
---

## Introduction

In this analysis, we simulate data to apply both **K-Means** and **K-Medoids** clustering algorithms, compare their results, and visualize the clustering outcomes.

We also measure the time taken for each clustering method.

## Load Required Libraries

```{r setup, message=FALSE, warning=FALSE}
# Install libraries if not already installed
# install.packages("cluster")
# install.packages("factoextra")
# install.packages("ggplot2")
# install.packages("tictoc")

# Load necessary libraries
library(cluster)      # For K-Medoids (PAM)
library(factoextra)   # For visualization
library(ggplot2)      # For plotting
library(tictoc)       # For measuring execution time
library(datasets)
```

```{r}

set.seed(42)

n_values <- seq(300, 1500, by = 200)

runtime_results <- data.frame(
  n_points = integer(),
  method = character(),
  time_seconds = numeric()
)

for (n in n_values) {
  cluster1 <- data.frame(x = rnorm(n/3, mean = 2, sd = 0.5), y = rnorm(n/3, mean = 2, sd = 0.5))
  cluster2 <- data.frame(x = rnorm(n/3, mean = 6, sd = 0.5), y = rnorm(n/3, mean = 6, sd = 0.5))
  cluster3 <- data.frame(x = rnorm(n/3, mean = 10, sd = 0.5), y = rnorm(n/3, mean = 2, sd = 0.5))
  
  data <- rbind(cluster1, cluster2, cluster3)
  colnames(data) <- c("feature1", "feature2")
  
  scaled_data <- scale(data)
  
  tic()
  kmeans_result <- kmeans(scaled_data, centers = 3, nstart = 10)
  kmeans_time <- toc(quiet = TRUE)
  
  tic()
  kmedoids_result <- pam(scaled_data, 3)
  kmedoids_time <- toc(quiet = TRUE)
  
  runtime_results <- rbind(
    runtime_results,
    data.frame(
      n_points = n,
      method = "K-Means",
      time_seconds = kmeans_time$toc - kmeans_time$tic
    ),
    data.frame(
      n_points = n,
      method = "K-Medoids",
      time_seconds = kmedoids_time$toc - kmedoids_time$tic
    )
  )
}
```

```{r}
ggplot(runtime_results, aes(x = n_points, y = time_seconds, color = method, group = method)) +
  geom_line(size = 1) +
  geom_point(size = 3) +
  labs(
    title = "Comparison: K-Means vs K-Medoids",
    x = "# of Data Points",
    y = "Execution Time (in seconds)",
    color = "Clustering Method"
  ) +
  theme_minimal() +
  scale_y_continuous(labels = scales::comma)
```

```{r data prep}
# Set random seed for reproducibility
set.seed(42)

# Create 3 clusters with normal distribution
n <- 300  # Total number of data points
cluster1 <- data.frame(x = rnorm(n/3, mean = 2, sd = 0.5), y = rnorm(n/3, mean = 2, sd = 0.5))
cluster2 <- data.frame(x = rnorm(n/3, mean = 6, sd = 0.5), y = rnorm(n/3, mean = 6, sd = 0.5))
cluster3 <- data.frame(x = rnorm(n/3, mean = 10, sd = 0.5), y = rnorm(n/3, mean = 2, sd = 0.5))

# Combine clusters into one dataset
data <- rbind(cluster1, cluster2, cluster3)
colnames(data) <- c("feature1", "feature2")

# Normalize the data
scaled_data <- scale(data)

```
# Apply K-Means Clustering

```{r k-means}
tic("K-Means")  # Start timer
kmeans_result <- kmeans(scaled_data, centers = 3, nstart = 10)
toc()  # Stop timer
```

# Apply K-Medoids Clustering
```{r k-medoids}
tic("K-Medoids")  # Start timer
kmedoids_result <- pam(scaled_data, 3)
toc()  # Stop timer

```

# Prep data for visualization
```{r get memberships }
data$KMeans_Cluster <- as.factor(kmeans_result$cluster)
data$KMedoids_Cluster <- as.factor(kmedoids_result$clustering)
```

# Plotting k-means clusters
```{r vis kmeans}
p1 <- ggplot(data, aes(x = feature1, y = feature2, color = 
KMeans_Cluster)) +
  geom_point(size = 3) +
  ggtitle("K-Means Clustering") +
  theme_minimal()

# Print the K-Means plot
print(p1)
```

# Plotting k-medoids clusters
```{r kmedoids}
p2 <- ggplot(data, aes(x = feature1, y = feature2, color = KMedoids_Cluster)) +
  geom_point(size = 3) +
  ggtitle("K-Medoids Clustering") +
  theme_minimal()

# Print the K-Medoids plot
print(p2)
```

```{r}
data(iris)

iris_data <- iris[, c("Sepal.Length", "Sepal.Width")]

iris_scaled <- scale(iris_data)

tic("K-Means (Iris)")
iris_kmeans <- kmeans(iris_scaled, centers = 3, nstart = 10)
kmeans_iris_time <- toc()

tic("K-Medoids (Iris)")
iris_kmedoids <- pam(iris_scaled, 3)
kmedoids_iris_time <- toc()

iris$KMeans_Cluster <- as.factor(iris_kmeans$cluster)
iris$KMedoids_Cluster <- as.factor(iris_kmedoids$clustering)
```

```{r}
p3 <- ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = KMeans_Cluster, shape = Species)) +
  geom_point(size = 3) +
  ggtitle("K-Means Clustering on Iris Dataset") +
  theme_minimal()
print(p3)
```

```{r}
p4 <- ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = KMedoids_Cluster, shape = Species)) +
  geom_point(size = 3) +
  ggtitle("K-Medoids Clustering on Iris Dataset") +
  theme_minimal()

print(p4)
```